String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-23 2023-1-19/sections/introduction.tex
revise bib
partially revised introduction

Java, JavaScript, and Python software use regex matching\cite{redos_fse2019}. Bounded repetition is widely used in regex matching\cite{regex_repeat} and is dangerous because Regular expression Denial of Service (ReDoS) may derive from it\cite{redos_lenka}. We have analyzed regular expressions sourced from the Internet\cite{regexlib}\cite{stackoverflow}, among which about 10\% include bounded repetition. Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications\cite{malware_detection_3_kudzu}). These statistics show that a resultful string solver reasoning about regular membership predicate with bounded repetition and \verb|length| operation is required. However, when the string constraints contain bounded repetition and lower bound on string variables, the state-of-art string solvers (\cite{z3seq}\cite{z3str3}\cite{z3str4}\cite{z3str3re}\cite{cvc5}\cite{ostrich}\cite{z3trau}\cite{trau}) lose efficiency. For example, CVC5\cite{cvc5} (which is the winner of QF\_Strings track in SMT-COMP 2022\cite{smt-comp}) failed to solve string constraints (\ref{eq:cvc5_timeout}) with timeout 60s.

@@

$\Sigma_{/a}\{1,300\}$ repeats a letter excluding $a$ from one time to 300 times. $\Sigma^*a\Sigma^*$ accepts a string containing $a$. It is easy to see that the string constraints are unsat. Nevertheless, CVC5 is hard to find a solution for it can not answer unsat until attempting all strings with lengths less than or equal to 300. The number of the searched string is exponential. Z3str3RE\cite{z3str3re} can not solve string constraints (\ref{eq:z3str3re_timeout}) because its length abstraction is over-approximation.

@@

On account of the large automaton corresponding to the regular expression, Ostrich(+)\cite{ostrich}\cite{atva2020} crashes on string constraints (\ref{eq:ostrich_timeout}) where $\Sigma_{[0-9A-Za-z]}$ is a digital letter or an English letter.
\begin{equation}\label{eq:ostrich_timeout}
  x\in \Sigma_{[0-9A-Za-z]}\{0, 63\}\wedge |x| > 0

@@

The rest of the paper is structured as follows: Section \ref{sec:pre} introduces the preliminaries. Section \ref{sec:overview} outlines the overall algorithm. Section \ref{sec:regex2acfa} illustrates how to construct an automaton from a regular expression. Section \ref{sec:algorithm} shows how to solve a string constraint with repetition. Section \ref{sec:implementation} presents the experimental results. Section \ref{sec:conclu} concludes the paper and discusses future work.

>>>>

Java, JavaScript, and Python software use regex matching\cite{redos_fse2019}. Bounded repetition is widely used in regex matching\cite{regex_repeat} and is dangerous because Regular expression Denial of Service (ReDoS) may derive from it\cite{redos_lenka}. We have analyzed regular expressions sourced from the Internet\cite{regexlib}\cite{stackoverflow}, among which about 10\% include bounded repetition. Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications\cite{malware_detection_3_kudzu}). These statistics show that a resultful string solver reasoning about regular membership predicate with bounded repetition and \verb|length| operation is required. However, such string constraints may lead to low performance. 
% the state-of-art string solvers (\cite{z3seq}\cite{z3str3}\cite{z3str4}\cite{z3str3re}\cite{cvc5}\cite{ostrich}\cite{z3trau}\cite{trau}) lose efficiency. 
For CVC5\cite{cvc5} (which is the winner of QF\_Strings track in SMT-COMP 2022\cite{smt-comp}), the string constraints are solved based on DPLL(T) structure. Each string constraint is seen as a literal boolean and assigned by true or false (Boolean abstraction). CVC5 repeatedly applies derivation rules on these assigned string constraints until the satisfiable solution is found or unsatisfiability is detected. The approach is fast and sound. Nevertheless, it is incomplete because it may not recognize unsatisfiable literals and may not terminate when unrolling regular expression. For example, CVC5 unrolls regular expression $\Sigma_{/a}\{1,300\}$ and $\Sigma^*a\Sigma^*$ when solving the string constraints \ref{eq:cvc5_timeout}. $\Sigma_{/a}\{1,300\}$ repeats a letter excluding $a$ from one time to 300 times. $\Sigma^*a\Sigma^*$ accepts a string containing $a$. It is easy to see that the string constraints are unsat.
However, CVC5 is hard to explore the unsatisfiability, for it can only answer once attempting all strings with lengths less than or equal to 300. The number of the searched string is exponential to the count. We test this instance on CVC5 and get a timeout result when the timeout is set to 60s.

@@

Z3Str3\cite{z3str3} is a string solver written on the famous SMT solver Z3\cite{z3}. It is based on the DPLL(T) structure and uses the same approach as CVC5, except that the derivation rules of Z3 differ from CVC5. Furthermore, Z3str3 exposes the structure of theory literals to the DPLL(T) SAT solver's branching heuristic, resulting in smarter decisions during its search. Z3Str3 can not solve string constraints (\ref{eq:cvc5_timeout}) as well because the optimization can not avoid the endless unwind of the regular expression. To handle the exponential search space of regular expression, automata-based methods were proposed. One of the automata-based string solver Ostrich\cite{ostrich}\cite{atva2020} encodes string constraints and regular expressions into automata and use the automata theory to solve them. It is sound and complete. But it is inefficient because it is hard to reduce the size of automata. When large counting is constrained on a complex sub-regex, like $ \backslash w \{1, 63\}$ where $\backslash w$ is an English character or a number character, the corresponding automaton in Ostrich becomes very large and can not be solved within 60s time limit. The reason is that Ostrich syntactically rewrites the counting operator to the union of finite repeatings of the sub-regex. For example, $\backslash w\{1, 63\}$ is rewritten to $\backslash w\mid\backslash w^2\cdots\mid\backslash w^{63}$ where $\backslash w^i$ means repeat $i\-th$ times on $\backslash w$. The automaton of $\backslash w$ contains $62$ states to accept each possible character, and the automaton of $\backslash w ^i$ contains $62*i$ states. The total number of states in the automaton is $62 + 62*2 + \cdots 62*63=124992$, causing the automaton to be unsolvable in practice. Trau\cite{z3trau}\cite{trau+} is another powerful automata-based solver. It relies on a Counter-Example Guided Abstraction Refinement (CEGAR) framework in which the under- and over-approximation of the automata are iteratively refined. Although it uses simple flat automata to represent regular expressions, it has to build unsolvable automata on regex $\backslash w\{1, 63\}$ as Ostrich because the solution of the flat automata should be within the solution of the original automata. 
Recently, Z3str3 has been extended to Z3str3RE\cite{z3str3re} by Berzish et al to handle string constants with regular memberships and length operation effectively. It uses automata to formalize the sematic of regular expression and applies several length-aware heuristics. Z3str3RE can avoid building unsolvable automaton because the length abstraction is directly derived from the regex, not the automaton. Unfortunately, the length abstraction is an over-approximation, so it can not solve string constraints (\ref{eq:z3str3re_timeout}) whose regular membership predicate contains complementary.

@@

@@

The rest of the paper is structured as follows: Section \ref{sec:pre} introduces the preliminaries. Section \ref{sec:overview} outlines the overall algorithm. Section \ref{sec:automaton} illustrates how to construct an automaton from a regular expression. Section \ref{sec:algorithm} shows how to solve a string constraint with repetition. Section \ref{sec:implementation} presents the experimental results. Section \ref{sec:conclu} concludes the paper and discusses future work.
